Extended Data

### 2021-10-13 doc/manuscript/Dow_Extended_Data.Rmd
answer some comments / add info to extended figure captions

growth_curves_all_caption  <- " **Extended Data Figure 2 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth.** Each line represents one year's growth for a given tree, fit with McMahon model. " #needs wordsmithing!!!

@@

pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective critical temperature window (CTW). For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands. (**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

tree_rings_TMIN_caption <- "Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 207 chronologies from 114 sites across eastern North America (Extended Data Figure 1). Chronologies are grouped by xylem porosity and ordered by mean April T_max. Chronology details are given in the Supplementary Information. (**NOTE: Figure still needs some work. Chronology numbers are off.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!! #info subject to change!!!

>>>>

growth_curves_all_caption  <- " **Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at our two focal study sites** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2) ..., with each line representing a year between 2000 and 2018. For stem growth, each line represents one year's growth for a given tree, fit with McMahon model. " #needs wordsmithing!!!

@@

pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Response of the timing of stem growth phenology to spring temperatures.** Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown. Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. 95% credible intervals are represented by bands. The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective CTW. For each CTW $T_{max}$. Modeled values from dendrometer band measurements are represented by dots. (**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF). Relationships between critical temperature window (CTW) $T_max$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown. Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. 95% credible intervals are represented by bands. (**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF). Relationships between critical temperature window (CTW) $T_min$ and 25%, 50%, and 75% total stem growth (first row), the length of the peak growing season ($L_pgs$; second row), maximum growth rate ($g_max$; third row), and total seasonal radial stem growth ($\Delta DBH$; fourth row) are shown. Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. 95% credible intervals are represented by bands. (**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

tree_rings_TMIN_caption <- "Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 208 chronologies from 114 sites across eastern North America (Extended Data Figure 1). Chronologies are grouped by xylem porosity and ordered by mean April T_max. Chronology details are given in the Supplementary Information. (**NOTE: Figure still needs some work. Chronology numbers are off.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!! #info subject to change!!!

### 2021-10-13 doc/manuscript/Dow_manuscript_Nature.Rmd
answer some comments / add info to extended figure captions

Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **146** <!--Cameron, please verify number--> trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **755** <!--Cameron, please verify number--> trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of leaf phenology (measured at ecosystem level via remote sensing).

@@

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p=0.001). 
Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest ~~when comparing the years with the warmest and coldest mean $T_{max}$ during the critical temperature window~~<!--Cameron, was this based on comparing just the warmest and coldest years, or a linear regression. We want the latter, and to remove the italic text -->, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
Specifically, at SCBI, RP growth advanced 1.9, 1.5, and 1.1 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 3.5, 3.5, and 3.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
At Harvard Forest, RP growth advanced 2.8, 5.0, and 7.2 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 7.9, 7.3, and 6.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.

@@

The second study site was Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.), located near the central Massachusetts town of Petersham.
Harvard forest is a secondary forest, having re-established following agricultural use *(time frame)* and significant hurricane damage in 1938. 
Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures  averaging -12° C in January and 19° C in July [@dorangeville_peak_2021].
<!-- Is this the right reference? -->

@@

Metal dendrometer bands were installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r  EDT_species`). 
Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 
The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).

@@

### Climate {-#Climate} 
For the dendrometer band analysis, we used climate data from local weather stations. 
Daily maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeo_2020]. 
Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
The R package climpact () was used to plot temperatures for visual inspection. 
Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. 
Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). 
(*how were gaps filled for Harvard? Or was the record good?*)

>>>>

Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **125** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **340** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of leaf phenology (measured at ecosystem level via remote sensing).

@@

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p=0.001). 
Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
Specifically, at SCBI, RP growth advanced 1.9, 1.5, and 1.1 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 3.5, 3.5, and 3.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.
At Harvard Forest, RP growth advanced 2.8, 5.0, and 7.2 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively, while DP growth advanced 7.9, 7.3, and 6.6 days/ $^\circ$C for $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively.

@@

The second study site was Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.), located near the central Massachusetts town of Petersham.
Harvard forest is a secondary forest, having re-established following agricultural use *(time frame)* and significant hurricane damage in 1938. 
Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures  averaging -12° C in January and 19° C in July [@dorangeville_peak_2021].
<!-- Is this the right reference? Yes, I got this info from the draft they sent. -->

@@

Metal dendrometer bands were installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r  EDT_species`). 
Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest. 
The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
Across years, the number of bands sampled averaged *~124* (range: 91-138) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).

@@

### Climate {-#Climate} 
For the dendrometer band analysis, we used climate data from local weather stations. 
Daily maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeo_2020]. 
Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
The R package climpact () was used to plot temperatures for visual inspection. 
Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. 
Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). 
(*how were gaps filled for Harvard? Or was the record good? --- Data came "pre-cleaned" as far as I could tell*)
